---
title: "Using missRanger"
date: "`r Sys.Date()`"
bibliography: "biblio.bib"
link-citations: true
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using missRanger}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
```

## Introduction

The aim of this vignette is to introduce the R package `missRanger` for imputation of missing values and to explain how to use it for multiple imputation.

`missRanger` uses the `ranger` package [@wright] to do fast missing value imputation by chained random forest. As such, it can be used as an alternative to `missForest`, a beautiful algorithm introduced in [@stekhoven]. Basically, each variable is imputed by predictions from a random forest using all other variables as covariables. `missRanger` iterates multiple times over all variables until the average out-of-bag prediction error of the models stops to improve.

Why should you consider `missRanger`?

- It is fast.

- It is flexible and intuitive to apply: E.g. calling `missRanger(data, . ~ 1)` would impute all variables univariately, `missRanger(data, Species ~ Sepal.Width)` would use `Sepal.Width` to impute `Species`.

- It can deal with most realistic variable types, even dates and times without destroying the original data structure.

- It combines random forest imputation with predictive mean matching. This generates realistic variability and avoids "new" values like 0.3334 in a 0-1 coded variable. Like this, `missRanger` can be used for realistic multiple imputation scenarios, see e.g. [@rubin] for the statistical background.

In the examples below, we will meet two functions from the `missRanger` package:

- `generateNA`: To replace values in a data set by missing values.

- `missRanger`: To impute missing values in a data frame.

## Installation

From CRAN:
```
install.packages("missRanger")
```

Latest version from github:
```
library(devtools)
install_github("mayer79/missRanger", subdir = "release/missRanger")
```

## Examples

We first generate a data set with about 20% missing values per column and fill them again by `missRanger`.

``` {r}
library(missRanger)
library(dplyr)

set.seed(84553)

head(iris)

# Generate data with missing values in all columns
irisWithNA <- generateNA(iris, p = 0.2)
head(irisWithNA)
 
# Impute missing values with missRanger
irisImputed <- missRanger(irisWithNA, num.trees = 100, verbose = 0)
head(irisImputed)
```

It worked! Unfortunately, the new values look somewhat unnatural due to different rounding. If we would like to avoid this, we just set the `pmm.k` argument to a positive number. All imputations done during the process are then combined with a predictive mean matching (PMM) step, leading to more natural imputations and improved distributional properties of the resulting values:

``` {r}
irisImputed <- missRanger(irisWithNA, pmm.k = 3, num.trees = 100, verbose = 0)
head(irisImputed)
```

Note that `missRanger` offers a `...` argument to pass options to `ranger`, e.g. `num.trees` or `min.node.size`. How would we use its "extra trees" variant with 50 trees?

``` {r}
irisImputed_et <- missRanger(
  irisWithNA, 
  pmm.k = 3, 
  splitrule = "extratrees", 
  num.trees = 50, 
  verbose = 0
)
head(irisImputed_et)
```

It is as simple!

Further note that `missRanger` does not rely on `tidyverse` but you can embed it into a `dplyr` pipeline (without `group_by`).

``` {r}
iris %>%
  generateNA() %>%
  missRanger(verbose = 0) %>%
  head()
```

By default `missRanger` uses all columns in the data set to impute all columns with missings. To override this behaviour, you can use an intuitive formula interface: The left hand side specifies the variables to be imputed (variable names separated by a `+`), while the right hand side lists the variables used for imputation.

``` {r}
# Impute all variables with all (default behaviour). Note that variables without
# missing values will be skipped from the left hand side of the formula.
m <- missRanger(irisWithNA, formula = . ~ ., 
                pmm.k = 3, num.trees = 10, seed = 1, verbose = 0)
head(m)

# Same
m <- missRanger(irisWithNA, pmm.k = 3, num.trees = 10, seed = 1, verbose = 0)
head(m)

# Impute all variables with all except Species
m <- missRanger(irisWithNA, . ~ . - Species, 
                pmm.k = 3, num.trees = 10, verbose = 0)
head(m)

# Impute Sepal.Width by Species 
m <- missRanger(irisWithNA, Sepal.Width ~ Species, 
                pmm.k = 3, num.trees = 10, verbose = 0)
head(m)

# No success. Why? Species contains missing values and thus can only 
# be used for imputation if it is being imputed as well
m <- missRanger(irisWithNA, Sepal.Width + Species ~ Species, 
                pmm.k = 3, num.trees = 10, verbose = 0)
head(m)

# Impute all variables univariatly
m <- missRanger(irisWithNA, . ~ 1, verbose = 0)
head(m)
```

## Imputation takes too much time. What can I do?

`missRanger` is based on iteratively fitting random forests for each variable with missing values. Since the underlying random forest implementation `ranger` uses 500 trees per default, a huge number of trees might be calculated. For larger data sets, the overall process can take very long.

Here are tweaks to make things faster:

- Use less trees, e.g. by setting `num.trees = 50`. Even one single tree might be sufficient. Typically, the number of iterations until convergence will increase with fewer trees though.

- Use smaller bootstrap samples by setting e.g. `sample.fraction = 0.1`.

- Use the less greedy `splitrule = "extratrees"`.

- Use a low tree depth `max.depth = 6`.

- Use large leafs, e.g. `min.node.size = 10000`.

- Use a low `max.iter`, e.g. 1 or 2.

### Examples evaluated on a normal laptop (not run here)

``` r
library(ggplot2) # for diamonds data
dim(diamonds) # 53940    10

diamonds_with_NA <- generateNA(diamonds)

# Takes 270 seconds (10 * 500 trees per iteration!)
system.time(
  m <- missRanger(diamonds_with_NA, pmm.k = 3)
)

# Takes 19 seconds
system.time(
  m <- missRanger(diamonds_with_NA, pmm.k = 3, num.trees = 50)
)

# Takes 7 seconds
system.time(
  m <- missRanger(diamonds_with_NA, pmm.k = 3, num.trees = 1)
)

# Takes 9 seconds
system.time(
  m <- missRanger(diamonds_with_NA, pmm.k = 3, num.trees = 50, 
    sample.fraction = 0.1)
)
```

## Trick: Use `case.weights` to weight down contribution of rows with many missings

Using the `case.weights` argument, you can pass case weights to the imputation models. This might be e.g. useful to weight down the contribution of rows with many missings.

### Example

``` {r}
# Count the number of non-missing values per row
non_miss <- rowSums(!is.na(irisWithNA))
table(non_miss)

# No weighting
m <- missRanger(irisWithNA, num.trees = 20, pmm.k = 3, seed = 5, verbose = 0)
head(m)

# Weighted by number of non-missing values per row. 
m <- missRanger(irisWithNA, num.trees = 20, pmm.k = 3, seed = 5, verbose = 0,
                case.weights = non_miss)
head(m)
```

## References
